Detection of Treatment-Subgroup Interactions in Clustered Datasets with Generalized Linear Mixed-effects Model Trees (GLMM trees)

### 2015-08-25 pkg/glmertree/paper/treatment_interactions.Rnw
started Sweave version of the manuscript
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@756 255f4249-2263-4417-b1ea-8498779e27a7

@@

\begin{figure}[!h]
    \tikzstyle{decision} = [diamond, draw]
    \tikzstyle{line} = [draw, -stealth, thick]
    \tikzstyle{elli}=[draw, ellipse, minimum height=8mm, text width=6em, text centered]
   \tikzstyle{block} = [draw, rectangle, text width=7em, text centered, minimum height=15mm, node distance=6em]
    \begin{tikzpicture}[paths/.style={->, thick, >=stealth'}]
        \node [elli] (node1) { };
        \node [elli, below of =node1, xshift=-9em] (node2) { };
        \node [elli, below of =node1, xshift=9em] (node3) { };
        \node [block, below of=node2, xshift=-4em, yshift=-1em] (block1){\small$\beta_{j0}=17.5$\\$\beta_{j1}=-5.0$\\$d_j=-1.0$};
        \node [block, below of=node2, xshift=4em, yshift=-1em] (block2){\small$\beta_{j0}=30.0$\\$\beta_{j1}=0.0$\\$d_j=0.0$};
        \node [block, below of=node3, xshift=-4em, yshift=-1em] (block3){\small$\beta_{j0}=30.0$\\$\beta_{j1}=0.0$\\$d_j=0.0$};
        \node [block, below of=node3, xshift=5em, yshift=-1em, text width=8em] (block4){\small$\beta_{j0}=42.5$\\$\beta_{j1}=5.0$\\$d_j=1.0$};
        \draw [paths] (node1) to node [above left]{\small$U_2\leq30$} (node2);
        \draw [paths] (node1) to node [above right]{\small$U_2>30$} (node3);
        \draw [paths] (node2) to node [ left]{\small$U_1\leq17$} (block1);
        \draw [paths] (node2) to node [ right]{\small$U_1>17$} (block2);
        \draw [paths] (node3) to node [ left]{\small$U_5\leq63$} (block3);
        \draw [paths] (node3) to node [ right]{\small$U_5>63$} (block4);
    \end{tikzpicture}
	\caption{Data-generating model for treatment-subgroup interactions. Parameter $d$ denotes the standardized mean difference between the outcomes of Treatment 1 and 2 (i.e., $\beta_{j1} / \sigma_{\epsilon}$).}
	\label{fig:modelC}
\end{figure}

@@

@@

@@

\begin{figure}[!ht]
    \includegraphics[width=12cm]{app_lmtree.pdf}
	\caption{GLM tree for prediction of posttreatment total scores on the Hamilton Rating Scale for Depression (HAM-D). The y-axes of the boxplots represent posttreatment HAM-D scores, and the x-axes represent treatment levels: cognitive behavior therapy (CBT) vs. pharmacotherapy (PHA).}
	\label{fig:lmtree_C&W}
\end{figure}

@@

\begin{figure}[!ht]
    \includegraphics[width=10cm]{app_lmertree.pdf}
	\caption{GLMM tree for prediction of posttreatment total scores on the Hamilton Rating Scale for Depression (HAM-D). The y-axes of the boxplots represent posttreatment HAM-D scores, and the x-axes represent treatment levels: cognitive behavior therapy (CBT) vs. pharmacotherapy (PHA).}
	\label{fig:lmertree_C&W}
\end{figure}

>>>>

%% for \usepackage{Sweave}
\SweaveOpts{echo=FALSE, results=hide}
\setkeys{Gin}{width=0.8\textwidth}

<<preliminaries>>=
library("foreign")
library("glmertree")
library("lattice")
@

@@

\begin{figure}[!h]
% \tikzstyle{decision} = [diamond, draw]
% \tikzstyle{line} = [draw, -stealth, thick]
% \tikzstyle{elli}=[draw, ellipse, minimum height=8mm, text width=6em, text centered]
% \tikzstyle{block} = [draw, rectangle, text width=7em, text centered, minimum height=15mm, node distance=6em]
% \begin{tikzpicture}[paths/.style={->, thick, >=stealth'}]
%     \node [elli] (node1) { };
%     \node [elli, below of =node1, xshift=-9em] (node2) { };
%     \node [elli, below of =node1, xshift=9em] (node3) { };
%     \node [block, below of=node2, xshift=-4em, yshift=-1em] (block1){\small$\beta_{j0}=17.5$\\$\beta_{j1}=-5.0$\\$d_j=-1.0$};
%     \node [block, below of=node2, xshift=4em, yshift=-1em] (block2){\small$\beta_{j0}=30.0$\\$\beta_{j1}=0.0$\\$d_j=0.0$};
%     \node [block, below of=node3, xshift=-4em, yshift=-1em] (block3){\small$\beta_{j0}=30.0$\\$\beta_{j1}=0.0$\\$d_j=0.0$};
%     \node [block, below of=node3, xshift=5em, yshift=-1em, text width=8em] (block4){\small$\beta_{j0}=42.5$\\$\beta_{j1}=5.0$\\$d_j=1.0$};
%     \draw [paths] (node1) to node [above left]{\small$U_2\leq30$} (node2);
%     \draw [paths] (node1) to node [above right]{\small$U_2>30$} (node3);
%     \draw [paths] (node2) to node [ left]{\small$U_1\leq17$} (block1);
%     \draw [paths] (node2) to node [ right]{\small$U_1>17$} (block2);
%     \draw [paths] (node3) to node [ left]{\small$U_5\leq63$} (block3);
%     \draw [paths] (node3) to node [ right]{\small$U_5>63$} (block4);
% \end{tikzpicture}
\setkeys{Gin}{width=0.8\textwidth}
<<dgp-tree, fig=TRUE, height=5, width=7>>=
fig4 <- party(
  partynode(1L,
    split = partysplit(2L, breaks = 30),
    kids = list(
      partynode(2L,
        split = partysplit(1L, breaks = 17),
        kids = list(
          partynode(3L, info = c(
	    expression(''),
	    expression(beta[j0] == '17.5'),
	    expression(''),
	    expression(beta[j1] == -'5.0'),
	    expression(''),
	    expression(d[j] == -'1.0')
	  )),
          partynode(4L, info = c(
	    expression(''),
	    expression(beta[j0] == '30.0'),
	    expression(''),
	    expression(beta[j1] == '0.0'),
	    expression(''),
	    expression(d[j] == '0.0')
	  )))),
      partynode(5L,
        split = partysplit(1L, breaks = 63),
        kids = list(
          partynode(6L, info = c(
	    expression(''),
	    expression(beta[j0] == '30.0'),
	    expression(''),
	    expression(beta[j1] == '0.0'),
	    expression(''),
	    expression(d[j] == '0.0')
	  )),
          partynode(7L, info = c(
	    expression(''),
	    expression(beta[j0] == '42.5'),
	    expression(''),
	    expression(beta[j1] == '5.0'),
	    expression(''),
	    expression(d[j] == '1.0')
	  )))))),
  data.frame(U1 = numeric(0), U2 = numeric(0))
)
plot(fig4, tp_args = list(FUN = identity, width = 9), tnex = 1.5)
@
\caption{Data-generating model for treatment-subgroup interactions. Parameter $d$ denotes the standardized mean difference between the outcomes of Treatment 1 and 2 (i.e., $\beta_{j1} / \sigma_{\epsilon}$).}
\label{fig:modelC}
\end{figure}

@@

<<depresseion-data>>=
metadata <- read.dta("Database IPDMA CBT PHA Version 11.dta")
metadata[metadata == 999] <- NA
metadata[metadata == 888] <- NA
vars <- c("studyid", "Tx_group", "Age", "Gender", "education", "ComorbidAnxietyDisorder", "HRSDt0", "HRSDt1")
factors <- c("studyid", "Tx_group", "Gender", "education", "ComorbidAnxietyDisorder")
metadata$education <- factor(metadata$education, ordered = TRUE)
for (i in 1:length(factors)) {metadata[, factors[i]] <- factor(metadata[,factors[i]])}
metadata <- metadata[vars] # select only relevant variables
metadata <- metadata[complete.cases(metadata[, vars]), ] # select only complete data
metadata <- metadata[!metadata$Tx_group == "placebo", ] # remove placebo observations
metadata$Tx_group <- factor(metadata$Tx_group)
@

@@

<<depression-trees>>=
formula <- HRSDt1 ~ Tx_group | Age + Gender + education + ComorbidAnxietyDisorder + HRSDt0

## GLM tree
lmtree_app <- lmtree(formula, data = metadata,
  offset = fitted(lm(HRSDt1 ~ HRSDt0), data = metadata))

## GLMM tree
lmertree_app <- lmertree(formula, HRSDt1 ~ (1 | studyid) + HRSDt0, data = metadata,
  initialRandomEffects = fitted(lm(HRSDt1 ~ HRSDt0, data = metadata)))
@

@@

\begin{figure}[!ht]
\setkeys{Gin}{width=\textwidth}
<<depression-glmtree, fig=TRUE, height=6, width=9>>=
plot(lmtree_app)
@
\caption{GLM tree for prediction of posttreatment total scores on the Hamilton Rating Scale for Depression (HAM-D). The y-axes of the boxplots represent posttreatment HAM-D scores, and the x-axes represent treatment levels: cognitive behavior therapy (CBT) vs. pharmacotherapy (PHA).}
\label{fig:lmtree_C&W}
\end{figure}

@@

\begin{figure}[!ht]
\setkeys{Gin}{width=0.75\textwidth}
<<depression-glmmtree, fig=TRUE, height=4, width=6>>=
plot(lmertree_app$Tree)
@
\caption{GLMM tree for prediction of posttreatment total scores on the Hamilton Rating Scale for Depression (HAM-D). The y-axes of the boxplots represent posttreatment HAM-D scores, and the x-axes represent treatment levels: cognitive behavior therapy (CBT) vs. pharmacotherapy (PHA).}
\label{fig:lmertree_C&W}
\end{figure}
